---
title: "Computational Statistics Assignment 1"
author: "Edwin Ngungi - 224082"
date: "2026-02-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question 1: Monte Carlo Integration
### 1.1 Estimation of Theta

```{r q1-estimation}
set.seed(42)

# Define the function to be integrated
f <- function(x) {
  x^2 * exp(-x)
}

# Parameters
n <- 10000  # Sample size
a <- 0      # Lower bound
b <- 2      # Upper bound

# Step 1: Generate random samples from Uniform(0, 2)
u <- runif(n, min = a, max = b)

# Step 2: Evaluate the function at these points
samples <- f(u)

# Step 3: Estimate theta
# Formula: (b - a) * average(f(u))
theta_hat <- (b - a) * mean(samples)
theta_hat
```
### 1.2 Monte Carlo Standard Error and Confidence Interval
```{r}
# Calculate values of f(u)
f_u <- f(u)

# Standard Error calculation

mc_se <- (b - a) * sd(f_u) / sqrt(n)

# 95% Confidence Interval
alpha <- 0.05
z_crit <- qnorm(1 - alpha/2)
ci_lower <- theta_hat - z_crit * mc_se
ci_upper <- theta_hat + z_crit * mc_se

cat("Monte Carlo SE:", mc_se, "\n")
cat("95% CI: [", ci_lower, ",", ci_upper, "]\n")
```
###1.3 Accuracy and Comparison
```{r}
# Exact value using R's numerical integration
exact_val <- integrate(f, 0, 2)$value

# Absolute error
abs_error <- abs(theta_hat - exact_val)

cat("Analytical Value:", exact_val, "\n")
cat("MC Estimate:", theta_hat, "\n")
cat("Absolute Error:", abs_error, "\n")
```
## Question 2: Exponential Distribution Simulation

### 2.1 Estimating P(2 < T < 5)
```{r q2-simulation}
set.seed(123)

# Parameters
n <- 10000
lambda <- 0.4

# Step 1: Generate lifetimes from Exponential(0.4)
lifetimes <- rexp(n, rate = lambda)

# Step 2: Identify lifetimes between 2 and 5 years
# This creates a logical vector (TRUE/FALSE)
successes <- (lifetimes > 2) & (lifetimes < 5)

# Step 3: Calculate the proportion
prob_estimate <- mean(successes)
prob_estimate
```
### 2.2 Stability of the Estimate
To observe how the estimate stabilizes, we repeat the simulation for sample sizes of 1,000, 5,000, and 10,000.
```{r}
set.seed(42)
lambda <- 0.4

# Define a function to calculate the estimate for a given n
estimate_prob <- function(n_size) {
  sim_data <- rexp(n_size, rate = lambda)
  mean(sim_data > 2 & sim_data < 5)
}

# Define the sample sizes requested in the assignment
sample_sizes <- c(1000, 5000, 10000)

# Run the simulation for each size
results <- sapply(sample_sizes, estimate_prob)

# Create a clean table to display the results
stability_table <- data.frame(
  n = sample_sizes,
  MC_Estimate = results
)

print(stability_table)
```
We observe that the estimates do not follow a linear trend (downward or upward), but rather fluctuate around a central value. As $n$ increases, the magnitude of these fluctuations decreases, leading to a more stable and accurate estimate.

### 2.3 Comparison with Theoretical Probability
The theoretical probability for an Exponential distribution $P(a < T < b)$ is calculated as $F(b) - F(a)$.
```{r}
# Define parameters
lambda <- 0.4
a <- 2
b <- 5

# Calculate theoretical probability using the CDF (pexp)
theoretical_prob <- pexp(b, rate = lambda) - pexp(a, rate = lambda)
print(theoretical_prob)
```
The theoretical probability is approximately 0.3139937. As the $n$ value increases, the Monte-Carlo estimate gets closer to the theoretical value.

## Question 3: Estimating Integrals over Infinite Limits

### 3.1 Naive Monte Carlo
We approximate the infinite limit with a large finite value $M=20$.
```{r q3-naive}
set.seed(42)
n <- 10000
M <- 20
u <- runif(n, 0, M)

# Estimate = (Width) * Average of f(u)
theta_naive <- M * mean(u * exp(-u))
theta_naive
```
### 3.2 Importance Sampling

```{r}
set.seed(42)
# 1. Generate samples from Exponential(1)
x_is <- rexp(n, rate = 1)

# 2. Calculate weights: f(x)/g(x) = (x * exp(-x)) / exp(-x) = x
weights <- x_is

# 3. The estimate is the mean of the weights
theta_is <- mean(weights)
theta_is
```
### 3.3 Comparison of Variance and Efficiency
```{r}
# 1. Variance of the Naive Estimator
# We must scale the variance by M^2 because theta_naive = M * mean(f(u))
var_naive <- M^2 * var(u * exp(-u)) / n

# 2. Variance of the Importance Sampling Estimator
var_is <- var(weights) / n

# 3. Efficiency Ratio

efficiency_ratio <- var_naive / var_is

cat("Variance (Naive):", var_naive, "\n")
cat("Variance (Importance Sampling):", var_is, "\n")
```
This is because Importance Sampling uses a proposal distribution that mimics the target function's behavior, concentrating samples in regions that contribute most to the integral. This reduces the variability of the individual estimates, leading to a more efficient and stable estimator.


# Question 4: Rare Event Estimation

### 4.1 Naive Monte Carlo Simulation
We attempt to estimate the probability $P(X > 4)$ for a standard normal variable $X \sim N(0, 1)$.

```{r}
set.seed(42)
n <- 10000

# Draw from Standard Normal
x_naive <- rnorm(n, mean = 0, sd = 1)

# Count how many are > 4
prob_naive <- mean(x_naive > 4)
prob_naive
```
### 4.2 Importance Sampling with a Shifted Normal Proposal
To estimate $P(X > 4)$ more accurately, we sample from a proposal distribution $g(x) \sim N(4, 1)$.
```{r}
set.seed(42)
n <- 10000

# 1. Generate samples from the Proposal Distribution N(4, 1)
x_is <- rnorm(n, mean = 4, sd = 1)

# 2. Calculate the weights: f(x) / g(x)
# f(x) = dnorm(x, 0, 1) [Target]
# g(x) = dnorm(x, 4, 1) [Proposal]
weights <- dnorm(x_is, mean = 0, sd = 1) / dnorm(x_is, mean = 4, sd = 1)

# 3. Estimate the probability
# We multiply the weight by an indicator (x > 4)
prob_is <- mean(weights * (x_is > 4))

cat("Importance Sampling Estimate:", prob_is)
```
### 4.3 Comparison and Discussion of Results
```{r}
# 1. THE TRUTH (for comparison)
theoretical_p <- pnorm(4, lower.tail = FALSE)

# 2. CALCULATE BIAS (How far are we from the truth?)
# Bias = Estimate - Truth
bias_naive <- prob_naive - theoretical_p
bias_is    <- prob_is - theoretical_p

# 3. CALCULATE VARIANCE (How "shaky" is the estimate?)
# For Naive: Variance of the TRUE/FALSE hits
var_naive <- var(x_naive > 4) / n

# For IS: Variance of the Weights
var_is    <- var(weights * (x_is > 4)) / n

# 4. RESULTS TABLE
summary_table <- data.frame(
  Metric   = c("Estimate", "Bias", "Variance"),
  Naive    = c(prob_naive, bias_naive, var_naive),
  Imp_Samp = c(prob_is, bias_is, var_is)
)

print(summary_table)
```
Importance Sampling is preferable because it achieves a level of precision that is not possible by Naive Monte Carlo at the same sample size. With a variance that is lower and a bias that is close to zero.

